---
title: "Using gatom package"
date: "2017-03-21"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using gatom package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r message=FALSE}
library(gatom)
library(data.table)
library(igraph)
```

First let's load data with atom mappings (`network` object),
enzyme annotations for mouse (`org.Mm.eg.gatom`)
and metabolite annotations (`met.kegg.db.rda`):

```{r}
data("networkEx")
data("org.Mm.eg.gatom.annoEx")
data("met.kegg.dbEx")
```


Loading input data:

```{r message=F}
data("met.de.rawEx")
data("gene.de.rawEx")

```

Getting atom graph

```{r}
g <- makeAtomGraph(network=networkEx, 
                   org.gatom.anno=org.Mm.eg.gatom.annoEx, 
                   gene.de=gene.de.rawEx,
                   met.db=met.kegg.dbEx, 
                   met.de=met.de.rawEx)
print(g)
```

Scoring graph

```{r}
gs <- scoreGraph(g, k.gene = 25, k.met=25)
```

Finding a module

```{r}
set.seed(42)
m <- solveSgmwcsRandHeur(gs, max.iterations = 2000)
```

```{r}
print(m)
E(m)$label
V(m)$label
```

We can save the module to dot format to generate pdf or svg file using `neato`
tool from graphviz suite.

```{r}
saveModuleToDot(m, file="M0.vs.M1.dot", name="M0.vs.M1")
system("neato -Tsvg M0.vs.M1.dot > M0.vs.M1.svg", ignore.stderr = T)
knitr::include_graphics("M0.vs.M1.svg")
```

The module can also be save to xgmml format and later opened in Cytoscape.

```{r}
saveModuleToXgmml(m, file="M0.vs.M1.xgmml", name="M0.vs.M1")
```


Next, reactions without highly changing genes but with high average 
expression can be added.

```{r}
m.ext <- addHighlyExpressedEdges(m, gs)
saveModuleToDot(m.ext, file="M0.vs.M1.ext.dot", name="M0.vs.M1")
system("neato -Tsvg M0.vs.M1.ext.dot > M0.vs.M1.ext.svg", ignore.stderr = T)
knitr::include_graphics("M0.vs.M1.ext.svg")
```


Sometimes, as in example above, the same metabolite can appear multiple 
times in the module via different atoms. In such cases it is useful
to either connect atoms belonging to the same metabolite with edges
with `connectAtomsInsideMetabolite` function.

```{r}
m1 <- connectAtomsInsideMetabolite(m.ext)
saveModuleToDot(m1, file="M0.vs.M1.v1.dot", name="M0.vs.M1")
system("neato -Tsvg M0.vs.M1.v1.dot > M0.vs.M1.v1.svg", ignore.stderr = T)
knitr::include_graphics("M0.vs.M1.v1.svg")
```

The other way is to collapse them into one vertex with `collapseAtomsIntoMetabolites`:

```{r}
m2 <- collapseAtomsIntoMetabolites(m.ext)
saveModuleToDot(m2, file="M0.vs.M1.v2.dot", name="M0.vs.M1")
system("neato -Tsvg M0.vs.M1.v2.dot > M0.vs.M1.v2.svg", ignore.stderr = T)
knitr::include_graphics("M0.vs.M1.v2.svg")
```

## Pre-generated annotations


```{r eval=FALSE}
load(url("http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/network.rda"))
load(url("http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/org.Mm.eg.gatom.anno.rda"))
load(url("http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/org.Mm.eg.gatom.anno.rda"))
load(url("http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/met.kegg.db.rda"))
```

Full dataset for example can be downloaded here:

```{r eval=FALSE}
library(readr)
met.de.raw <- read_tsv("http://artyomovlab.wustl.edu/publications/supp_materials/GAM/Ctrl.vs.MandLPSandIFNg.met.de.tsv.gz")


gene.de.raw <- read_tsv("http://artyomovlab.wustl.edu/publications/supp_materials/GAM/Ctrl.vs.MandLPSandIFNg.gene.de.tsv.gz")

```
