---
title: "Using gatom package"
author: "Anstasiia Gainullina, Mariia Emelianova, Alexey Sergushichev"
date: "October 2023"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: false
    self_contained: true
vignette: >
  %\VignetteIndexEntry{Using gatom package}
  %\VignetteEngine{knitr::rmarkdown}
  \VignetteEncoding{UTF-8}"
---

This tutorial describes an R-package for finding active metabolic modules 
based on high throughput data. The pipeline takes as input transcriptional and/or metabolic data 
and finds a metabolic subnetwork (module) most regulated between the two 
conditions of interest. 

The package relies on the active module analysis framework
developed in [BioNet package](https://bioconductor.org/packages/BioNet),
but extends it to work with metabolic reaction networks. 
Further, it illustrates the usage of [mwcsr package](https://cran.r-project.org/package=mwcsr)
which provides a number of solvers for Maximum Weight Connected Subgraph problem and its variants.

Example of using the pipeline include:

* studying metabolic differences between pro- and anti-inflammatory macrophage activation ([Jha et al, 2015](http://dx.doi.org/10.1016/j.immuni.2015.02.005));
* studying metabolic rewiring associated with glucose-independent tumor growth ([Vinent at al, 2015](http://dx.doi.org/10.1016/j.molcel.2015.08.013));
* identification of deregulation of energy metabolism in Trem2-deficient macrophages ([Ulland et al, 2017](http://doi.org/10.1016/j.cell.2017.07.023));
* identification of inositol-triphosphate metabolism  activation in monocytes in fasting mice ([Jordan et al, 2019](https://doi.org/10.1016/j.cell.2019.07.050)).

More details on the pipeline are available in [Sergushichev et al, 2016](http://dx.doi.org/10.1093/nar/gkw266) and [Emelianova et al, 2022](https://doi.org/10.1093/nar/gkac427).

# Installation

You can install **gatom** via `BiocManager`:
```{R eval = FALSE}
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("gatom")
```


# Example workfow

In this example we will find an active metabolic module based on macrophage 
activation gene expression and metabolomics data 
([Jha et al, 2015](http://dx.doi.org/10.1016/j.immuni.2015.02.005)).
For improved performance here we will consider a simplified version of the data.
See [Example on full data and full network](#example-full) section for the real-scale analysis.


```{r message=FALSE}
library(gatom)
library(data.table)
library(igraph)
library(mwcsr)
```

First let's load the example tables with input differential gene expression 
and metabolite abundance data for LPS-IFNg stimulated macrophages compared to controls:

```{r message=FALSE}
data("gene.de.rawEx")
print(head(gene.de.rawEx))
data("met.de.rawEx")
print(head(met.de.rawEx))
```

Next we will load example network related objects: global reaction network (`networkEx` object),
metabolite annotations (`met.kegg.dbEx`), and organism-specific enzyme annotations for mouse (`org.Mm.eg.gatom.annoEx`).

```{r}
data("networkEx")
data("met.kegg.dbEx")
data("org.Mm.eg.gatom.annoEx")
```

Here `networkEx` object contain information about `r nrow(networkEx$reactions)` KEGG reactions, their atom mappings and relation to enzymes:

```{r}
str(networkEx, max.level=1, give.attr = FALSE)
```

Object `met.kegg.dbEx` contains information about `r nrow(met.kegg.dbEx$metabolites)` KEGG metabolites, including mappings from HMDB and ChEBI:

```{r}
str(met.kegg.dbEx, max.level=2, give.attr = FALSE)
```

Object `org.Mm.eg.gatom.annoEx` contains mouse-specific mapping between enzyme classes
and genes, as well as mapping between different types of gene identifiers:

```{r}
str(org.Mm.eg.gatom.annoEx, max.level=2, give.attr = FALSE)
```

Then we create a metabolic graph with atom topology from the loaded data. 
Depending on `topology` parameter, the graph vertices can correspond either 
to `atoms` or `metabolites`. For metabolite topology, 
see [Using metabolite-level network](#met-topology) section.

```{r}
g <- makeMetabolicGraph(network=networkEx, 
                        topology="atoms",
                        org.gatom.anno=org.Mm.eg.gatom.annoEx, 
                        gene.de=gene.de.rawEx,
                        met.db=met.kegg.dbEx, 
                        met.de=met.de.rawEx)
print(g)
```

After creating the metabolic graph, we then score it, obtaining an instance of 
Signal Generalized Maximum Weight Subgraph (SGMWCS) problem.

The size of the module can be controlled by changing scoring parameters `k.gene`
and `k.met`. The higher the values of scoring parameters are, the bigger the
module is going to be. 

```{r message=FALSE, warning=FALSE}
gs <- scoreGraph(g, k.gene = 25, k.met = 25)
```

Then we initialize an SMGWCS solver. Here, we use a heuristic relax-and-cut 
solver `rnc_solver` for simplicity.

See `mwcsr` package documentation for more solver options, or 
[Using exact solver](#exact-solver) section for the recommended way.

```{r}
solver <- rnc_solver()
```

Then we find an active metabolic module with chosen solver and scored graph:

```{r message=FALSE, warning=FALSE}
set.seed(42)
res <- solve_mwcsp(solver, gs)
m <- res$graph
```

The result module is an `igraph` object that captures the most regulated
reactions:

```{r}
print(m)
head(E(m)$label)
head(V(m)$label)
```

The module can be plotted in R Viewer with `createShinyCyJSWidget()`. 
Here, red color corresponds to up-regulation (positive log-2 fold change) and 
green to down-regulation (negative log-2 fold change). Blue nodes and edges 
come from data with absent log-2 fold change values. Bigger size of nodes and 
width of edges reflect lower p-values.

```{r}
createShinyCyJSWidget(m)
```

# Saving modules

We can save the module to graphml format with `write_graph()` function from `igraph`:

```{r}
write_graph(m, file = file.path(tempdir(), "M0.vs.M1.graphml"), format = "graphml")
```

Or it can be saved to an interactive html widget:

```{r message=FALSE}
saveModuleToHtml(module = m, file = file.path(tempdir(), "M0.vs.M1.html"), 
                 name="M0.vs.M1")
```

We can also save the module to dot format:

```{r}
saveModuleToDot(m, file = file.path(tempdir(), "M0.vs.M1.dot"), name = "M0.vs.M1")
```

Such dot file can be further used to generate svg file using `neato` tool 
from graphviz suite if it is installed on the system:

```{r eval=FALSE}
system(paste0("neato -Tsvg ", file.path(tempdir(), "M0.vs.M1.dot"),
              " > ", file.path(tempdir(), "M0.vs.M1.svg")), 
       ignore.stderr=TRUE)
```

Alternatively, the module can be saved to pdf format with a nice layout. 

You may vary the meaning of repel force and the number of iterations of 
repel algorithm for label layout. Note, that the larger your graph is the softer 
force you should use. 

You may also set different seed for different variants of edge layout with `set.seed()`.

```{r results="hide", message=FALSE, warning=FALSE}
set.seed(42)
saveModuleToPdf(m, file = file.path(tempdir(), "M0.vs.M1.pdf"), name = "M0.vs.M1", 
                n_iter=100, force=1e-5)
```


# Example on full data and full network {#example-full}

Let's now look at how the analysis will work with the full dataset and the full network.
For this case we will be using the combined network instead of KEGG network 
(see [Networks](#networks) for details on the network types).

```{r message=FALSE}
library(R.utils)
library(data.table)
```

The full macrophage LPS+IFNG-activation dataset can be downloaded from 
[http://artyomovlab.wustl.edu/publications/supp_materials/GAM/](http://artyomovlab.wustl.edu/publications/supp_materials/GAM/):

```{r message=FALSE}
met.de.raw <- fread("http://artyomovlab.wustl.edu/publications/supp_materials/GAM/Ctrl.vs.MandLPSandIFNg.met.de.tsv.gz")
gene.de.raw <- fread("http://artyomovlab.wustl.edu/publications/supp_materials/GAM/Ctrl.vs.MandLPSandIFNg.gene.de.tsv.gz")
```

Full pre-generated combined network, corresponding metabolite annotation, 
and enzyme annotation can be downloaded from  [http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/](http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/):

```{r}
network.combined <- readRDS(url("http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/network.combined.rds"))
met.combined.db <- readRDS(url("http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/met.combined.db.rds"))

org.Mm.eg.gatom.anno <- readRDS(url("http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/org.Mm.eg.gatom.anno.rds"))
```

For better work of the combined network we highly recommend using additional 
supplementary gene files (see [Supplementary Genes](#suppl-genes)).

```{r}
gene2reaction.extra <- fread("http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/gene2reaction.combined.mmu.eg.tsv", colClasses="character")
```

Running `gatom` on the combined network and the full dataset:

```{r}
cg <- makeMetabolicGraph(network=network.combined,
                         topology="atoms",
                         org.gatom.anno=org.Mm.eg.gatom.anno,
                         gene.de=gene.de.raw,
                         met.db=met.combined.db,
                         met.de=met.de.raw,
                         gene2reaction.extra=gene2reaction.extra)

cgs <- scoreGraph(cg, k.gene = 50, k.met = 50)

solver <- rnc_solver()
set.seed(42)
sol <- solve_mwcsp(solver, cgs)
cm <- sol$graph
cm
```

The result module for combined network:

```{r message=FALSE, warning=FALSE}
createShinyCyJSWidget(cm)
```


# Networks {#networks}

We provide four types of networks that can be used for analysis:

1. KEGG network
2. Rhea network
3. Combined network
4. Rhea lipid subnetwork 

## KEGG {#kegg-network}

KEGG network consists of `network.kegg.rds` & `met.kegg.db.rds` files and is 
based on [KEGG database](https://www.genome.jp/kegg/kegg1.html). 

Both metabolites and reactions in KEGG network have KEGG IDs.

This network was generated with the pipeline available 
[here](https://github.com/ctlab/KEGG-network-pipeline).
For extra details on KEGG network you can also reference 
[shinyGatom](https://doi.org/10.1093/nar/gkac427) and 
[GAM](https://doi.org/10.1093/nar/gkw266) articles.

```{r}
network <- readRDS(url("http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/network.kegg.rds"))
met.db <- readRDS(url("http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/met.kegg.db.rds"))
```

Running `gatom` with KEGG network on full dataset:

```{r}
kg <- makeMetabolicGraph(network=network, 
                        topology="atoms",
                        org.gatom.anno=org.Mm.eg.gatom.anno, 
                        gene.de=gene.de.raw,
                        met.db=met.db, 
                        met.de=met.de.raw)

kgs <- scoreGraph(kg, k.gene = 50, k.met = 50)

solver <- rnc_solver()
set.seed(42)
sol <- solve_mwcsp(solver, kgs) 
km <- sol$graph
km
```

```{r message=FALSE, warning=FALSE}
createShinyCyJSWidget(km)
```

## Rhea

Rhea network consists of `network.rhea.rds` & `met.rhea.db.rds` files and 
is based on [Rhea database](https://www.rhea-db.org/). 

Reactions in Rhea have their own IDs, but unlike KEGG, metabolite IDs come 
from a separate database -- [ChEBI database](https://www.ebi.ac.uk/chebi/).

This network was generated with the pipeline available 
[here](https://github.com/ctlab/Rhea-network-pipeline).
For extra details on Rhea network you can also reference 
[shinyGatom](https://doi.org/10.1093/nar/gkac427) article.

To use Rhea network download the following files:

```{r}
network.rhea <- readRDS(url("http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/network.rhea.rds"))
met.rhea.db <- readRDS(url("http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/met.rhea.db.rds"))
```

For proper work of the Rhea network we also need a corresponding 
supplementary gene file (ref. [Supplementary Genes](#suppl-genes)).

```{r}
gene2reaction.extra <- fread("http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/gene2reaction.rhea.mmu.eg.tsv", colClasses="character")
```

And run `gatom` on Rhea network:

```{r}
rg <- makeMetabolicGraph(network=network.rhea,
                         topology="atoms",
                         org.gatom.anno=org.Mm.eg.gatom.anno,
                         gene.de=gene.de.raw,
                         met.db=met.rhea.db,
                         met.de=met.de.raw,
                         gene2reaction.extra=gene2reaction.extra)

rgs <- scoreGraph(rg, k.gene = 50, k.met = 50)

solver <- rnc_solver()
set.seed(42)
sol <- solve_mwcsp(solver, rgs)
rm <- sol$graph
rm
```

Result Rhea network module:

```{r message=FALSE, warning=FALSE}
createShinyCyJSWidget(rm)
```

## Combined network

Combined network comprises not only KEGG and Rhea reactions, but also transport 
reactions from [BIGG database](http://bigg.ucsd.edu/).

This means that reactions in such network have either KEGG or Rhea or BIGG IDs, 
and metabolite IDs are KEGGs and ChEBIs.

## Rhea lipid subnetwork

Rhea lipid subnetwork is subset of Rhea reactions that involve lipids, 
and it consists of `network.rhea.lipids.rds` & `met.rhea.lipids.db.rds` files. 

This network was generated with the pipeline available 
[here](https://github.com/ctlab/Rhea-network-pipeline).
For extra details on Rhea lipid subnetwork you can also reference 
[shinyGatom](https://doi.org/10.1093/nar/gkac427) article.

```{r}
network.lipids <- readRDS(url("http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/network.rhea.lipids.rds"))
met.lipids.db <- readRDS(url("http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/met.lipids.db.rds"))
```

For proper work of the lipid network we will also need a corresponding 
supplementary gene file (ref. [Supplementary Genes](#suppl-genes))

```{r}
gene2reaction.extra <- fread("http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/gene2reaction.rhea.mmu.eg.tsv", colClasses="character")
```

To test lipid network we will use example lipidomics data for WT mice control 
vs high fat diet comparison from 
[ST001289 dataset](https://www.metabolomicsworkbench.org/data/DRCCMetadata.php?Mode=Study&StudyID=ST001289).

```{r}
met.de.lipids <- fread("https://artyomovlab.wustl.edu/publications/supp_materials/GATOM/Ctrl.vs.HighFat.lipid.de.csv")
```

For lipid network we recommend setting topology to `metabolites` 
(ref. [Using metabolite-level network](#met-topology)):

```{r}
lg <- makeMetabolicGraph(network=network.lipids,
                         topology="metabolites",
                         org.gatom.anno=org.Mm.eg.gatom.anno,
                         gene.de=NULL,
                         met.db=met.lipids.db,
                         met.de=met.de.lipids,
                         gene2reaction.extra=gene2reaction.extra)

lgs <- scoreGraph(lg, k.gene = NULL, k.met = 50)

solver <- rnc_solver()
set.seed(42)
sol <- solve_mwcsp(solver, lgs)
lm <- sol$graph
lm
```

Result lipid subnetwork module:

```{r message=FALSE, warning=FALSE}
createShinyCyJSWidget(lm)
```

If IDs for metabolite differential abundance data are of type `Species` we can 
process metabolite labels into more readable ones:

```{r message=FALSE, warning=FALSE}
lm1 <- abbreviateLabels(lm, orig.names = TRUE, abbrev.names = TRUE)

createShinyCyJSWidget(lm1)
```


# Misc

## Supplementary gene files {#suppl-genes}

For combined, Rhea and lipid networks we provide supplementary files with genes 
that either come from proteome or are not linked to a specific enzyme. 
These files are organism-specific and are also available at  [http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/](http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/).

```{r}
network.combined <- readRDS(url("http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/network.combined.rds"))
met.combined.db <- readRDS(url("http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/met.combined.db.rds"))
gene2reaction.extra <- fread("http://artyomovlab.wustl.edu/publications/supp_materials/GATOM/gene2reaction.combined.mmu.eg.tsv", colClasses="character")

gg <- makeMetabolicGraph(network=network.combined, 
                         topology="atoms",
                         org.gatom.anno=org.Mm.eg.gatom.anno, 
                         gene.de=gene.de.raw,
                         met.db=met.combined.db, 
                         met.de=met.de.raw, 
                         gene2reaction.extra=gene2reaction.extra)
gg
```

## Non-enzymatic reactions

Optionally, we can also preserve non-enzymatic reactions that are found in the network. 
This can be done by setting `keepReactionsWithoutEnzymes` to `TRUE`:

```{r}
ge <- makeMetabolicGraph(network=network.combined, 
                         topology="atoms",
                         org.gatom.anno=org.Mm.eg.gatom.anno, 
                         gene.de=gene.de.raw,
                         met.db=met.combined.db, 
                         met.de=met.de.raw, 
                         gene2reaction.extra=gene2reaction.extra, 
                         keepReactionsWithoutEnzymes=TRUE)
ge
```

## Using exact solver {#exact-solver}

For proper analysis quality we recommend to use exact SGMWCS solver `virgo_solver()`, 
which requires Java (version >= 11) and CPLEX (version >= 12.7) to be installed. 
If the requirements are met you can then find a module as following:

```{r eval=FALSE}
vsolver <- virgo_solver(cplex_dir=Sys.getenv("CPLEX_HOME"), 
                        threads=4, penalty=0.001, log=1)
sol <- solve_mwcsp(vsolver, gs) 
m <- sol$graph
```

Edge penalty option there is used to remove excessive redundancy in genes.

## Running with no metabolite data

If there is no metabolite data in your experiment assign `met.de` and `k.met` to `NULL`:

```{r}
g <- makeMetabolicGraph(network=networkEx, 
                        topology="atoms",
                        org.gatom.anno=org.Mm.eg.gatom.annoEx, 
                        gene.de=gene.de.rawEx,
                        met.db=met.kegg.dbEx, 
                        met.de=NULL)
gs <- scoreGraph(g, k.gene = 50, k.met = NULL)
```

## Running with no gene data

If there is no gene data in your experiment assign `gene.de` and `k.gene` to `NULL`:

```{r}
g <- makeMetabolicGraph(network=networkEx, 
                        topology="atoms",
                        org.gatom.anno=org.Mm.eg.gatom.annoEx, 
                        gene.de=NULL,
                        met.db=met.kegg.dbEx, 
                        met.de=met.de.rawEx)
gs <- scoreGraph(g, k.gene = NULL, k.met = 50)
```


## Using metabolite-level network {#met-topology}

Sometimes it could make sense to work with metabolite-metabolite topology 
of the network, not atom-atom one. Such network is less structured, but contains 
more genes.

```{r}
gm <- makeMetabolicGraph(network=network, 
                        topology="metabolite",
                        org.gatom.anno=org.Mm.eg.gatom.anno, 
                        gene.de=gene.de.raw,
                        met.db=met.db, 
                        met.de=met.de.raw)

gms <- scoreGraph(gm, k.gene = 50, k.met = 50)

solver <- rnc_solver()
set.seed(42)
sol <- solve_mwcsp(solver, gms) 
mm <- sol$graph
mm
```

## Pathway annotation

To get functional annotation of obtained modules by KEGG and Reactome
metabolic pathways, we can use hypergeometric test with `fora()` function from `fgsea` package.

```{r}
foraRes <- fgsea::fora(pathways=org.Mm.eg.gatom.anno$pathways,
                       genes=E(km)$gene,
                       universe=unique(E(kg)$gene),
                       minSize=5)
foraRes[padj < 0.05]
```

Optionally, redundancy in pathways can be decreased with `collapsePathwaysORA()` function:

```{r}
mainPathways <- fgsea::collapsePathwaysORA(
  foraRes[padj < 0.05],
  pathways=org.Mm.eg.gatom.anno$pathways,
  genes=E(km)$gene,
  universe=unique(E(kg)$gene))
foraRes[pathway %in% mainPathways$mainPathways]
```


```{r}
sessionInfo()
```
